---
title: "Finding Indicators for Stock Prices"
author: "Eric Seltzer"
date: "2024-02-09"
categories: [finance, code, analysis]
---

```{r}
#| echo: false
#| warning: false
#| message: false
#| output: false
library(tidyverse)
library(tidyquant)
library(ggrepel)
library(here)
theme_set(theme_minimal())
companies <- read_csv(here("data/companies.csv"))
prices <- read_csv(here("data/prices.csv"))
```


Introduction:

After finding a data set showing the stock prices of big tech companies, I decided I wanted to investigate to see if there were any correlations to find trends in how their prices perform. This dataset began in 2010 and went to January of 2023 so I decided to cut down the time frame to the most recent five years. I also chose to use the six companies in this dataset with the highest observed closing prices. This allows for less clutter in my visualizations.

```{r}
#| warning: false
pricesNew <-
  prices |>
  filter(date > '2018-01-01')

prices |>
  group_by(stock_symbol) |>
  summarise(max = max(close)) |>
  arrange(desc(max)) |>
  head(n = 6)

pricesNew <- 
  pricesNew|>
  group_by(stock_symbol) |>
  filter(stock_symbol == c("NFLX", "ADBE", "TSLA", "MSFT", "NVDA", "META"))
```

# Plot 1
```{r}
pricesNew |>
  ggplot(aes(x = date,
             y = close)) +
  geom_line(aes(color = stock_symbol),
            linewidth = .9) +
  scale_color_brewer(palette = "Paired") + 
  labs(x = "Date",
       y = "Closing Price ($)",
       title = "Closing Stock Prices from 6 Big Tech Companies, 2018-2023",
       color = "Tickers",
       caption = "(based on data from Yahoo Finance via Kaggle)") +
  scale_x_date(date_minor_breaks = "6 months",
               date_breaks = "1 year",
               date_labels = "%b %y") +
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1,
                                   vjust = 1),
        plot.title = element_text(size = 16,
                                  hjust = 0),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 14))
```

My first plot shows the change in price for these six companies and it is easy to see that around January 2022, they all began to take a hit in their price. This begged the question, were there any indicators that this was going to happen? In order to attack this, I decided to investigate the short- and long-term moving averages for each stock, specifically the exponential moving average. An exponential moving average calculates the average of the prices using a weighting multiplier that assigns more weight to later data. In general, when a short-term EMA crosses a long-term EMA, this signifies either a price increase or decrease.

To calculate this, I had to make a new data set with the dates extending further back in order to get more observations for the moving averages. I also calculated the average price so it can be used as the y-variable.

```{r}
#| output: FALSE
pricesMA <-
  prices |>
  group_by(stock_symbol) |>
  filter(stock_symbol == c("NFLX", "ADBE", "TSLA", "MSFT", "NVDA", "META")) |>
  pivot_longer(cols = 3:6,
               values_to = "price",
               names_to = "market")

pricesMA <-
  pricesMA |>
  group_by(stock_symbol, date) |>
  summarise(avgPrice = mean(price))

prices_select <-
  prices |>
  filter(stock_symbol == c("NFLX", "ADBE", "TSLA", "MSFT", "NVDA", "META"))
```

# Plot 2
```{r}
#| warning: false
pricesMA |>
  ggplot(aes(x = date,
             y = avgPrice)) +
  geom_ma(ma_fun = EMA,
          n = 50,
          aes(color = "50 EMA"))+
  geom_ma(ma_fun = EMA,
          n = 200,
          aes(color = "200 EMA")) +
  facet_wrap(vars(stock_symbol),
             scales = "free_y") +
    scale_x_date(date_breaks = "4 year",
                 date_labels = "%b %y") +
  labs(y = "Dollars ($)",
       colour = "Moving Averages",
       title = "Short- and Long-Term EMA Compared to Closing Prices, 2010-2023",
       caption = "(based on data from Yahoo Finance via Kaggle)/nEMAs are calculated from library(tidyquant)") +
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1,
                                   vjust = 1),
        plot.title = element_text(size = 16,
                                  hjust = 0),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 14),
        panel.spacing = unit(1, "lines"),
        strip.text.x = element_text(color = "black", 
                                    face = "bold"),
        strip.background = element_rect(colour = "black",
                                        fill = "white"),
        legend.title = element_blank()) +
  geom_line(data = prices_select,
            aes(x = date,
                y = close),
            linetype = 1,
            color = "black",
            alpha = .2)
```

Conclusion:

In this plot, I plotted the moving averages against the closing prices from Plot 1 except this time I faceted by each company in order to compare the companies individually. This leads us to our answer to the original question of if there is a trend or indicator of the price decrease. Typically, when a short-term EMA crosses above the long-term EMA, this is a signal that price might continue to move up. On the opposite hand, this means that if short-term EMA crosses below long-term EMA, price might continue to move down. In every company shown, up until the closing price drop, the short-term was above the long-term. Once as the gap between them closed, and short-term came closer to long-term, the closing prices began to decrease. In some of the companies, the short-term even dipped below the long term. In conclusion, using moving averages can be a good indicator for these companies prices going down, but likely isn't the root of the scenario. That can be left to many other more complex factors.